Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} develops Thurstone's theory to
address these challenges. The development integrates core theoretical
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
discusses the implications of the findings and explores avenues for
future research. Finally, Section~\ref{sec-conclusion} summarizes the
study's key insights.

@@

\section{Developing Thurstone's theory}\label{sec-theoretical}

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed development. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this,
researcher design a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli), to
produce a dichotomous outcome that reflects the relative quality of each
text (trait). Based on this setup, researchers reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Finally, following current CJ practices, researchers
then rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by developing Thurstone's theory using
the structural approach to causal inference. Specifically, it leverages
the capabilities of this approach to formalize the combination of the
core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to enhancing statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

@@

In line with Thurstone's theory, the other variables present in
SCM~\ref{fig-cj03_scm} include the texts' discriminal processes
\((T_{ha}, T_{ib})\) and their discriminal difference \((D_{hiabjk})\)
(see Section~\ref{sec-thurstone_theory}). Additionally, based on the
arguments presented in Section~\ref{sec-theory-issue1b} and the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
the SCM~\ref{fig-cj03_scm} incorporates the judges' biases \((B_{kj})\).
Together with the outcome, these variables form the preliminary set of
endogenous variables,
\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\caption{\label{fig-cj03}Conceptual-population model, preliminary
integration of theoretical principles.}

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, showing the direction of causal
influence. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
for simplicity, following the \emph{standard representation}. However,
this section adopts the \emph{magnified representation} of DAGS, as
including exogenous variables can enhance clarity and reveal potential
issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

Figure~\ref{fig-cj03_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. This graph illustrates the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and the practical aspects of CJ experiments. It also
distinguishes between observable endogenous variables, such as the
outcome, and variables considered latent within the CJ framework, such
as the text's discriminal processes, their discriminal difference, and
the judges' biases.

Although specifying a system's data structure is not mandatory for using
SCMs and DAGs, defining one can sometimes enhance clarity and facilitate
the description of the system. Thus, to simplify the description of the
CJ system, we first define the vectors \(I\) and \(J\), along with the
matrices \(IA\) and \(JK\), as follows:

@@

Readers can better visualize the structure of these vectors and matrices
by considering small values for the parameters corresponding to number
of individuals \((n_{I} = 5)\), the number of available stimuli per
individual \((n_{A} = 2)\), the total number of judges \((n_{J} = 3)\),
and the number of judgments each judge performs \((n_{K} = 3)\). Using
these values, the matrices described in equations (\ref{eq-mat12}) and
(\ref{eq-mat3}) take the following form:

@@

Thus, using the data structure defined by the matrices in equations
(\ref{eq-mat12}) and (\ref{eq-mat3}), we can now re-express
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
vectorized form, as shown in Figure~\ref{fig-cj04}. In this
representation, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
judges' biases \(B_{JK}\) are no longer scalar values but vectors. These
vectors capture all the values from the conceptual population.
Specifically, \(O^{cp}_{RPUV}\) and \(D_{RPUV}\) are vectors of length
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
\(T_{IA}\) and \(B_{JK}\) are vectors of length \((n_{I} \cdot n_{A})\)
and \((n_{J} \cdot n_{K})\), respectively.

@@

\caption{\label{fig-cj04}Conceptual-population model, vectorized form of
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag}.}

@@

Building on the arguments developed in Section~\ref{sec-theory-issue2}
and the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model includes two hierarchical structural
components. Each component defines how different observed or latent
variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition facilitates analyses that
integrate the trait measurement and hypothesis testing processes,
simultaneously accounting for the traits and their associated
uncertainties, thereby reducing bias and enhancing the measurement
precision.

Figure~\ref{fig-cj09} displays the two hierarchical structural
components for the motivating example. Specifically, the top branch of
DAG~\ref{fig-cj09_dag} depicts the first component. This component
describes how the students' writing-quality trait \(T_{I}\) and a set of
\emph{relevant} text-related variables \(X_{IA}\) causally influence the
texts' written-quality trait \(T_{IA}\). The DAG also highlights the
idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
trait that neither the students' trait nor other relevant text-related
variables can explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors
of length \(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively.
Additionally, \(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.

Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} student-related variables \(X_{I}\) causally
influence the students' writing-quality trait \(T_{I}\). The DAG also
includes the idiosyncratic errors \(e_{I}\), which capture variations in
the the students' traits that the relevant student-related variables
cannot explain. Here, \(e_{I}\) is a latent vector of length \(n_{I}\),
and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(p_{I}\)
columns.

@@

\caption{\label{fig-cj09}Conceptual-population model, full integration
of theoretical principles.}

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component. This component describes how the judges' bias
\(B_{J}\) and a set of \emph{relevant} judgment-related variables
\(Z_{JK}\) causally influence the biases associated with each text
\(B_{JK}\). The DAG also highlights the idiosyncratic errors \(e_{JK}\),
which capture variations in the texts' bias that neither the judges'
bias nor other relevant judgment-related variables can explain. In this
description, \(B_{J}\) and \(e_{JK}\) are latent vectors of length
\(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively. Additionally,
\(Z_{JK}\) is an observed matrix with a dimension of
\((n_{J} \cdot n_{K})\) rows and \(p_{JK}\) columns.

Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} judge-related variables \(Z_{J}\) causally
influence the judges' bias \(B_{J}\). The DAG also accounts for
idiosyncratic errors \(e_{J}\), capturing the variations in the judges'
bias that relevant judge-related variables cannot explain. Here,
\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
observed matrix with \(n_{J}\) rows and \(p_{J}\) columns. Notably, all
variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} constitute the set of endogenous variables \(V\)
and structural equations \(F\) for the conceptual-population model.

In summary,

@@

A cautious reader will notice that the model descriptions of
Section~\ref{sec-theory-theoretical_P} rely on several relevant
variables. In this study, a \emph{relevant} or a \emph{sufficient
adjustment} set of variables (potentially an empty set) refers to those
that satisfy the \emph{backdoor criterion} \citep[pp.~37]{Neal_2020},
that is, variables that block all non-causal paths without introducing
new ones \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.

Incorporating these variables into the conceptual-population model
enables researchers to identify different estimands of interest and
their causal effects, even with observational data. This identification
works because conditioning on these variables ensures that the estimator
meets several critical properties, including confounding elimination
\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
the statistical analysis to a subset of the population defined by the
values of a given variable \citep[pp.~32]{Neal_2020}.

Naturally, the validity of causal claims now hinges on the assumption
that these variables serve as a sufficient adjustment set, a property
that cannot be guaranteed for any variable. Nevertheless, as
\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
real world from observed data inevitably requires assumptions, and these
requirements hold true for both observational and experimental data.

>>>>

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's theory to
address these challenges. The extension integrates core theoretical
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
discusses the implications of the findings and explores avenues for
future research. Finally, Section~\ref{sec-conclusion} summarizes the
study's key insights.

@@

\section{Extending Thurstone's theory}\label{sec-theoretical}

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed extension. In this example, researchers aim to determine: ``To
what extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, researcher
design a CJ experiment by randomly assigning students (individuals) to
two groups, each receiving a different teaching method. Judges then
compare pairs of students' written texts (stimuli), to produce a
dichotomous outcome that reflects the relative quality of each text
(trait). Based on this setup, researchers reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Finally, following current CJ practices, researchers
then rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's theory using
the structural approach to causal inference. Specifically, it leverages
the capabilities of this approach to formalize the combination of the
core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to enhancing statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, based on the arguments in Section~\ref{sec-theory-issue1b}
and the recommendations of \citet{Andrich_1978} and
\citet{Wainer_et_al_1978}, the SCM incorporates the judges' biases
\((B_{kj})\). Together with the outcome, these variables form the
preliminary set of endogenous variables,
\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\caption{\label{fig-cj03}Conceptual-population model, scalar form.}

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, showing the direction of causal
influence. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
for simplicity, following a \emph{standard representation}. However,
this section adopts the \emph{magnified representation} of DAGs, as
including exogenous variables in their representation can enhance
clarity and reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
variables based on Thurstone's theory and the practical aspects of CJ
experiments. Particularly, the graph shows how the texts' discriminal
processes \((T_{ha}, T_{ib})\) influence their discriminal difference
\((D_{hiabjk})\), which in turn determines the outcome. It also depicts
the judges' biases \((B_{kj})\) as an additional factor influencing the
texts' discriminal difference. Furthermore, the DAG helps distinguish
between observed endogenous variables, such as the outcome, and latent
endogenous variables, which include the texts' discriminal processes,
their discriminal difference, and the judges' biases.

Although specifying a system's data structure is not mandatory for using
SCMs and DAGs, defining one can sometimes enhance clarity and facilitate
the description of the system. Thus, to simplify the description of a CJ
experiment, we first define the vectors \(I\) and \(J\), along with the
matrices \(IA\) and \(JK\), as follows:

@@

Readers can better visualize the structure of these vectors and matrices
by considering small values for the parameters corresponding to number
of individuals, the number of available stimuli per individual, the
total number of judges, and the number of judgments each judge performs.
For instance, assuming \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
\(n_{K} = 3\), the matrices described in equations (\ref{eq-mat12}) and
(\ref{eq-mat3}) take the following form:

@@

Thus, using the data structure defined by the matrices in equations
(\ref{eq-mat12}) and (\ref{eq-mat3}), we can re-express
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
vectorized form, as shown in Figure~\ref{fig-cj04}. In this depiction,
the outcome \(O^{cp}_{RPUV}\), the texts' discriminal difference
\(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are represented as vectors rather than scalar values.
These vectors encompass all values from the conceptual population.
Specifically, \(O^{cp}_{RPUV}\) is an observed vector of length
\({n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
\(D_{RPUV}\) is a latent vector of the same dimension. Additionally,
\(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\), respectively.

@@

\caption{\label{fig-cj04}Conceptual-population model, vectorized form.}

@@

Moreover, building on the arguments developed in
Section~\ref{sec-theory-issue2} and the principles of Structural
Equation Modeling (SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item
Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model includes two hierarchical structural
components. Each component defines how different observed or latent
variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition facilitates analyses
integrating the trait measurement and hypothesis testing processes,
simultaneously accounting for the traits and their associated
uncertainties, thereby reducing bias and enhancing measurement
precision.

Figure~\ref{fig-cj09} displays the two hierarchical structural
components for the motivating example. Specifically, the top branch of
DAG~\ref{fig-cj09_dag} depicts the first component. This component
describes how the students' writing-quality trait \((T_{I})\) and a set
of \emph{relevant} text-related variables \((X_{IA})\) causally
influence the texts' written-quality trait \((T_{IA})\). The DAG also
highlights the idiosyncratic errors \((e_{IA})\), which capture
variations in the texts' trait that neither the students' trait nor
other relevant text-related variables can explain. Here, \(T_{I}\) and
\(e_{IA}\) are latent vectors of length \(n_{I}\) and
\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
\(q_{IA}\) columns.

Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} student-related variables \((X_{I})\) causally
influence the students' writing-quality trait \((T_{I})\). The DAG also
includes the idiosyncratic errors \((e_{I})\), which capture variations
in the students' traits that the relevant student-related variables
cannot explain. Here, \(e_{I}\) is a latent vector of length \(n_{I}\),
and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
columns.

@@

\caption{\label{fig-cj09}Conceptual-population model, final vectorized
form.}

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component. This component describes how the judges' bias
\((B_{J})\) and a set of \emph{relevant} judgment-related variables
\((Z_{JK})\) causally influence the biases associated with each text
\(B_{JK}\). The DAG also highlights the idiosyncratic errors
\((e_{JK})\), which capture variations in the texts' bias that neither
the judges' bias nor other relevant judgment-related variables can
explain. In this description, \(B_{J}\) and \(e_{JK}\) are latent
vectors of length \(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively.
Additionally, \(Z_{JK}\) is an observed matrix with a dimension of
\((n_{J} \cdot n_{K})\) rows and \(q_{JK}\) columns.

Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} judge-related variables \((Z_{J})\) causally
influence the judges' bias \((B_{J})\). The DAG also accounts for
idiosyncratic errors \((e_{J})\), capturing the variations in the
judges' bias that relevant judge-related variables cannot explain. Here,
\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
observed matrix with \(n_{J}\) rows and \(q_{J}\) columns. Notably, all
variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} constitute the set of endogenous variables \(V\)
and structural equations \(F\) for the conceptual-population model.

Overall, the conceptual-population model extends Thurstone's theory by
introducing key innovations to address the limitations discussed in
Section~\ref{sec-theory-issues}. These enhancements include accounting
for judges' biases and integrating hierarchical structural components.
Nevertheless, despite its statistical and measurement advantages, the
model still relies on the unrealistic assumption that researchers can
access data capturing all repeated judgments from every judge for each
pair of stimuli produced by every individual in the population. Since
this assumption is rarely met in practice, a more realistic scenario
must be considered.

@@
